Statistical Inference for inter-arrival times of extreme events in bursty time series

### 2020-08-01 article_springer/article_springer.Rmd
fix typos

We assume a dataset of the form $\{(t_i, x_i), i \in I\}$,
where $t_i$ are timestamps and $x_i$ are magnitudes.
In this article, we focus on modelling the distribution of threshold
*exceedance times* rather than the *exceedances* themselves. For the latter, 
see the POT approach [@smith1984threshold;
@leadbetter1991basis; @davison1990models; @beirlantBook Ch.5.3;
@embrechts2013modelling Ch. 6.5.; @ColesBook Ch.4.].
With the previous section setting the stage, we assume that only the large
magnitudes (aka "shocks") follow a MRP.  The smaller magnitudes are possibly
nuisance data that may be irrelevant with respect to the modelling of extremes
and their waiting times. 
Accordingly, we assume a threshold $\ell$, and discard all
data where $x_i \le \ell$, yielding the thresholded dataset
$\{(t^{(\ell)}_i, x^{(\ell)}_i), i \in I(\ell)\}$, with $x^{(\ell)}_i=x_i-\ell$, where $I(\ell):=\left\{i: x_i \le \ell \right\} \subseteq I$.

@@

: We approximated $p_{\ell}$, the probability that an event is larger than $l$, by its relative frequency. One could also approximate this tail probability by the GPD distribution fitted to the exceedances.

@@

We plot the tail parameter estimates $\hat \beta(k)$ against $k$ 
for each of the $m=100$ simulation runs. 
Thin grey lines represent individual simulation runs, 
and the thicker black line is their mean. 
Recall that $k$ is the index of the order statistics of $J_k$ at which the 
threshold $\ell$ is placed. 
<!-- Note that the Hill estimator is based on the waiting times, not on the  -->
<!-- inter-exceedance times, meaning that the estimation is based on more data.  -->
<!-- If one were to base the Hill-estimation only on the inter-exceedance times,  -->
<!-- the sample size would be too small and it would be impossible to deduce an  -->
<!-- estimate for the tail parameter from the plots (compare Figure \ref{fig:Hillplots}).  -->

<!-- In the right-hand Hill-estimator column, the $k$ on the $x$-axis corresponds to -->
<!-- $r$ from \eqref{eq:Hill}. -->
<!-- The Hill estimator appears to be nicely unbiased on Pareto data, -->
<!-- but compared to the other two estimators, it has a higher variance,  -->
<!-- as can be seen from the wider spread of the simulated curves.  -->
<!-- This high variance means that that a single estimate is less reliable compared -->
<!-- to the other two estimators.  -->
<!-- This effect is amplified if one only has a thresholded, and thus smaller dataset -->
<!-- available, to the extent that the Hill estimates become severely flawed -->
<!-- (see Section \ref{sec:ML}).  -->

@@

<!-- Since there is no Hill-estimate of a scale parameter, there are only two columns.  -->

@@

iid :

@@

The fit with a Mittag-Leffler distribution ($\beta = 0.9$) seems to be good (see Figure \ref{fig:flare-diagnostics-3}), though there are signs that the power-law tail tapers off for very large inter-threshold crossing times. There is no apparent dependence between threshold exceedance times and event magnitudes seen in the copula plot (see Figure \ref{fig:flare-diagnostics-2}). We also conduct a bootstrapped LRT for the null hypothesis of exponentially distributed inter-arrival times and received a $p$-value of $p<0.01$.

@@

```
The closer $\beta$ is to $1$, the more the hazard rate mimics that of an 
exponential distribution (a constant function, see Figure \ref{fig:hazard}, 
right panel).

@@

When we introduced the CTRE model, we assumed that all events are iid This assumption is likely sufficient but not necessary for our limit theorem to hold. Moreover, any data below a (minimum) threshold $\ell_0$ is discarded for CTREs, and hence need not satisfy the iid assumption. For the purposes of statistical inference, we merely require that the inter-threshold-crossing times are iid

The CTRE approach to model "non-Poissonian" threshold crossing times should be contrasted with the well-documented approach of clusters of extremes, see e.g. @ferro2003inference. When the underlying stochastic process is *stationary*, the exceedances of high thresholds form, asymptotically, a *Cluster Poisson Process*.
In the setting of this article, however, clustering-like dynamics occur due to the *non-stationarity* of the underlying renewal-reward process, which has infinite-mean renewal times. 
Hence CTRE and Cluster Poisson Process shouldn't be viewed as competing methods, as the underlying data generating processes are quite different. To differentiate between the two models we use the term *bursty* which is standard in the context of heavy-tailed inter-arrival times in the physics community [e.g. @Barabasi2005;@Vasquez2006;@Karsai2012;@Vajna2013]). In weighing the evidence for either of the two data generating processes, criteria could be developed, based e.g. on measures of surprise [@Lee15], which may prove to be valuable for future applied statisticians.

>>>>

We assume a dataset of the form $\{(t_i, x_i), i \in I\}$,
where $t_i$ are timestamps and $x_i$ are magnitudes.
In this article, we focus on modelling the distribution of threshold
*exceedance times* rather than the *exceedances* themselves. For the latter, 
see the POT approach [@smith1984threshold;
@leadbetter1991basis; @davison1990models; @beirlantBook Ch.5.3;
@embrechts2013modelling Ch. 6.5.; @ColesBook Ch.4.].
With the previous section setting the stage, we assume that only the large
magnitudes (or "shocks") follow a MRP.  The smaller magnitudes are possibly
nuisance data that may be irrelevant with respect to the modelling of extremes
and their waiting times. 
Accordingly, we assume a threshold $\ell$, and discard all
data where $x_i \le \ell$, yielding the thresholded dataset
$\{(t^{(\ell)}_i, x^{(\ell)}_i): i \in I(\ell)\}$, with $x^{(\ell)}_i=x_i-\ell$, where $I(\ell):=\left\{i \in I: x_i \ge \ell \right\} \subseteq I$.

@@

: We approximated $p_{\ell}$, the probability that an event is larger than $l$, by its relative frequency. One could also approximate this tail probability via the GPD distribution fitted to the exceedances.

@@

We plot the tail parameter estimates $\hat \beta(k)$ against $k$ 
for each of the $m=100$ simulation runs. 
Thin grey lines represent individual simulation runs, 
and the thicker black line is their mean. 
Recall that $k$ is the index of the order statistics of $J_k$ at which the 
threshold $\ell$ is placed.

@@

@@

iid:

@@

The fit with a Mittag-Leffler distribution ($\beta = 0.9$) seems to be good (see Figure \ref{fig:flare-diagnostics-3}), although there are signs that the power-law tail tapers off for very large inter-threshold crossing times. There is no apparent dependence between threshold exceedance times and event magnitudes seen in the copula plot (see Figure \ref{fig:flare-diagnostics-2}). We also conduct a bootstrapped LRT for the null hypothesis of exponentially distributed inter-arrival times and received a $p$-value of $p<0.01$.

@@

The closer $\beta$ is to $1$, the more the hazard rate mimics that of an 
exponential distribution (a constant function, see Figure \ref{fig:hazard}, 
right panel).

@@

When we introduced the CTRE model, we assumed that all events are iid. This assumption is likely sufficient but not necessary for our limit theorem to hold. Moreover, any data below a (minimum) threshold $\ell_0$ is discarded for CTREs, and hence need not satisfy the iid assumption. For the purposes of statistical inference, we merely require that the IETs are iid.

The CTRE approach to model "non-Poissonian" threshold crossing times should be contrasted with the well-documented approach of clusters of extremes, see e.g. @ferro2003inference. When the underlying stochastic process is *stationary*, the exceedances of high thresholds form, asymptotically, a *Cluster Poisson Process*.
In the setting of this article, however, clustering-like dynamics occur due to the *non-stationarity* of the underlying renewal-reward process, which has infinite-mean renewal times. 
Hence CTRE and Cluster Poisson Process should not be viewed as competing methods, as the underlying data generating processes are quite different. To differentiate between the two models we use the term *bursty* which is standard in the context of heavy-tailed inter-arrival times in the physics community [e.g. @Barabasi2005;@Vasquez2006;@Karsai2012;@Vajna2013]). In weighing the evidence for either of the two data generating processes, criteria could be developed, based e.g. on measures of surprise [@Lee15], which may prove to be valuable for future applied statisticians.
